---
title: "BIO331 -- Lab 07: Bioinformatics and Working with Strings of Biological Data"
subtitle: "Bioinformatics"
author: Dee Ruttenberg
date: Mar. 18, 2026
output: html_document
---

Today we're going to apply our work on R to Bioinformatic data sets. The typical output of a DNA sequencing experiment is a fastq file, consisting of often millions of configs which need to be stitched together (do you remember the two ways we do that?). Fastq files are different from fasta files as they contain quality scores -- this is useful to see how reliable your fastq data is.  The first tool we are going to use is Fastqc, to interpret the results of our fastq files.

# Quality Control

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Rqc")
```

We will use a specific small dataset used specifically for teaching bioinformatics. Rqc will output an .html file (essentially a web page).  For this class, I will ask that you understand specifically the results of 'Average Quality', 'Cycle-Specific Average Quality', 'Cycle-specific Quality Distribution', and 'Cycle-Specific Base Call Proportion'.

```{r}
library(Rqc)
folder <- system.file(package="ShortRead", "extdata/E-MTAB-1147")
rqc(path = folder, pattern = ".fastq.gz")
```

# Bioconductor

Bioconductor and Biostrings are especially powerful tools to examine fasta files in R.  In this section, we'll learn some tools to manipulate them.  

```{r}
library(Biostrings)
seqs <- readDNAStringSet("Full_cmv_259sequences_Aln.fasta")
```

As Fasta files are labeled, you can identify specific sequences, either by loading a specfic label or identifying a keyword:

```{r}
my_seq <- seqs["Europe_Czech Republic_KY490069.1_Human_betaherpesvirus_5_strain_PRA7_complete_genome"]
names_with_uk <- grep("United Kingdom", names(seqs), ignore.case = TRUE)
UK_seqs <- seqs[names_with_uk]
UK_subset <- subseq(UK_seqs, start = 20000, end = 25000)
```

You can, using R, code an aligned genome as a matrix, in order to identify snps. 

```{r}
mat <- as.matrix(UK_seqs)
snp_positions <- which(apply(mat, 2, function(col) {
  bases <- col[col %in% c("A","C","G","T")]
  length(unique(bases)) > 1
}))
snp_matrix <- mat[, snp_positions]
snp_list <- lapply(snp_positions, function(pos) {
  data.frame(
    Sequence = rownames(mat),
    Position = pos,
    Base = mat[, pos]
  )
})
names(snp_list) <- paste0("Pos_", snp_positions)
```

#Alignment

The final skill we need to learn is how to align .fasta files to a reference genome.  The data we have today is aligned so we'll have to arbitrarily "unalign" it.  We will use our first sequence as the reference.  

```{r}
UK_seqs_nogap <- DNAStringSet(gsub("-", "", as.character(UK_seqs))) 
ref <- seqs[6]
```

The region of interest from hcmv will be our query site.   

```{r}
library(Biostrings)
query <- readDNAStringSet("example.fasta")
ref <- seqs["Europe_Germany_JX512204.1_Human_herpesvirus_5_strain_HAN16_complete_genome"]
query <- DNAString(toupper(as.character(query)))
ref <- DNAStringSet(gsub("-", "", as.character(ref))) 

hits <- matchPattern(query, ref[[1]], max.mismatch=20)

```

We can now look at our aligned data!

```{r}
hits
```

# Homework Question 1: Interpretting QC data.

a. Look at the 'Cycle-specific Base Call Proportion' rqc file.  What is the ratio of A/C/G/T in cycles 1-10. What is the ratio of A/C/G/T in cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the cycles have a different ratio.
#It seems that the ratio's are very even in the first 10 cycles, but AT ratios are lower than the CG ratio in cycles 65-75. I believe that C(Cytosine) and G(Gaunine) are typically paired together in DNA, so that might be why there are two different paired ratios

b. Look at the 'Cycle specific Quality Distribution'. What is the typical quality of cycles 1-10?  What is the typical quality of cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the cycles have a different quality
#The quality is very high for cycles 1-10, while in cycles 65-75, it is much lower quality

# Homework Question 2: Loading a File, Reverse Complements

For one of the research projects, we will be studying a specific protein coding site of hcmv associated with the transition of hcmv from latent to active. Align the following two queries to any genome in the reference file. Which one successfully aligns? What does that mean about the site of interest?


```{r}
query <- readDNAStringSet("hcmv15-hcmv1b-1.fasta")
reversecompquery <- reverseComplement(query)
query <- DNAString(toupper(as.character(query)))
reversecompquery <- DNAString(toupper(as.character(reversecompquery)))

ref <- seqs["Europe_Germany_JX512204.1_Human_herpesvirus_5_strain_HAN16_complete_genome"]
ref <- DNAStringSet(gsub("-", "", as.character(ref))) 

ALIGN1 <- matchPattern(query, ref[[1]], max.mismatch=20)
ALIGN2 <- matchPattern(reversecompquery, ref[[1]], max.mismatch=20)

#reverse complement aligns to the reference at site 187075 to 191699. The site of interest is reversed compared to the rest of the genome? DNA revering mutation in this area possibly?
```

# Homework Question 3: Isolating a Site

Isolate the 'hcmv15-hcmv1b-1.fasta' site in all 259 seqs. Your final result should be a DNAString set of 259 sequences about 4-5 kb long.  
#Basically, show the string in ALIGN2 in ALL sequences, not just the one reference

```{r}
query <- readDNAStringSet("hcmv15-hcmv1b-1.fasta")
reversecompquery <- reverseComplement(query)
query <- DNAString(toupper(as.character(query)))
reversecompquery <- DNAString(toupper(as.character(reversecompquery)))

ref <- seqs["hcmv15-hcmv1b-1.fasta"]
ref <- DNAStringSet(gsub("-", "", as.character(ref))) 


test <- matchPattern(reversecompquery, ref[[1]], max.mismatch=20)
```
